Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-09-08 fstats_pca_lewontin.tex
revision figures

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}. In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species. As such, population genetic theory is essential to understand how this works.

@@

\subsection{Formal Definition of PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, all results are valid if we treat each individual as its own population. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains the most important variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	\begin{equation}
	\MY = \MC\MX = (\mathbf{U} \MSINGULAR) \mathbf{V}^T = \MP\ML\text{,}
	\end{equation}

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure. The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often used to look for outliers, which might be indicative of selection \citep[e.g][]{francois2010}.

	There are a number of choices when constructing a PCA from genetic data, for example SNPs are sometimes standardized by the inverse of their standard deviations \citep{patterson2006}, or low-quality samples are projected onto PCs created from high-quality samples \citep{patterson2006}. In order to simplify the relationship between PCA and $F$-statistics, I assume no standardization and no missing data.

\subsection{Formal Definition to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

 Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
	\end{align}
\end{subequations}

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}. The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}. This interpretation is useful to understand a number of applications. The outgroup-$F_3$-statistic $F_3(O; U, X_i)$, for example, is useful if we have an unknown population $U$, and want to find its closest relatives from a panel of populations $X_i$. The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The population $X_i$ with the largest value is the most closely related population out of the reference sample. The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

Most commonly however, $F_3$ and $F_4$ are used as admixture tests \citep{patterson2012}: Negative values of  $F_3(X_1; X_2, X_3) < 0$ correspond to a branch with negative genetic drift, which is a violation of treeness. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}.

To move away from trees and graph models, I build upon the geometric framework of \cite{oteo-garcia2021}. Here, we think of each population as a point in the data space $\mathbb{R}^S$, made up of the allele frequency at each SNP. Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between two populations $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (dot) product between these two vectors. These dot products are useful for a variety of projections that use population structure.

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the approximation error of all $F_2$-statistics.

@@

\subsection{When are admixture-$F_3$ statistics negative?}
Given two source populations $X_1$, $X_2$, can we predict which populations $X_x$ could be considered admixed between these populations based on PCA? Since the allele frequencies of $X_x$ are intermediate between those of $X_1$ and $X_2$, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \citep{brisbin2012, mcvean2009}.

The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. the set of points $X_x$ for which 
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}\text{.}
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$ (The overline notation denotes a line segment). $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).

\paragraph{$F_3$ on a 2D PCA-plot.} 
 If we project this $n$-ball on a two-dimensional plot, $\overline{X_1X_2}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot. This geometry is perhaps easiest visualized on a globe. If we look at the globe from a view point parallel to the equator, both the north and south poles are visible at the very edge of the circle. But if we look at it from above the north pole, the north- and south-poles will be at the very same point.

@@

\paragraph{Example}
As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g. Sardinian, Finnish) have, in fact, positive $F_3$-values. This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure. This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.  Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle. This allows us to immediately conclude that their $F_3$-statistics must be positive.

@@

\subsection{$F$-statistics as projections}
The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of a vector onto another one. This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.  The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$. This can be useful for interpretation: if potential source pulations are related according to a tree, the admixture axis will be orthogonal to the tree branches connecting potential source populations, and the population with the most negative $F$-statistic will be the most likely source \citep{patterson2012}.

On the other hand, if population relationships are more gradual, the interpretation is different: The lowest $F_3$-statistics will correspond to the populations where the $n$-sphere is biggest, i.e. the ends of the cline.  This is a practical issuel for example \cite{patterson2012} observed that many Western Eurasian populations have the most negative $F_3$-statistics if one of the source population is from Europe, and one from South America, which can be thought of a gradient in relationships throughout Eurasia and the Americas. Depending on whether we think of population relationships as more clinal or tree-like, negative $F_3$-statistics will thus have very different interpretations.

This projection argument also helps to interpret Outgroup-$F_3$-statistic on a PCA-plot (Figure \ref{fig:geom}D). In this context, we interpret the statistic $F_3(O; U, X_i)$ as the projection of $X_i$ onto the line segment $\overline{OU}$ from the outgroup $O$ to the test population $U$. The population $X_i$ with the largest $F_3$-statistic will have the largest projected segment, and is thus interpreted as the population most closely related to $U$.

>>>>

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\newcommand{\vectorproj}[2][]{\textit{proj}_{#1}#2}

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}. In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species. As such, population genetic theory is essential to understand how this works.

@@

\subsection{Formal Definition to $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci. Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

The three $F$-statistics can then be defined as
\begin{subequations}
	\begin{align}
	F_2(X_1, X_2) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	\\
	F_3(X_1; X_2, X_3) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) \\	
	F_4(X_1, X_2; X_3, X_4) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) 	\text{,}
	\end{align}
\end{subequations}
The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
	2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
	2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
	\end{align}
\end{subequations}

As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

Thinking of $F$-statistics as branch lengths is useful to understand a number of applications: In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}. One way to do that is using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ denotes an outgroup, and the $X_i$ are a panel of populations that are candidates for the closest match. The  highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The intuition is given in Figure \ref{fig:outgroupf3}A; where the outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ is highlighted. It represents the length of the branch from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the statistic. In contrast to a simple genetic distance, the sample time has no effect: The branch between $X_U$ and $X_2$, would be shorter than to one between $X_U$ and $X_1$, but the path to the shared junction and hence the $F_3$-statistic would be the same. Larger sets of $F_3$ and $F_4$-statistics are also frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}. The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact equations can be found in ][]{peter2016}. While admixture graphs are commonly considered, other alternative models exist. In \citep{peter2016}, I showed that $F_4$ is sensitive to many non-symmetric population structure models. $F_3$ is more robust, but may also be negative if substantial population substructure exists.

\subsubsection{Geometric interpretation of $F$-statistics}
For very large data sets, it is hard to justify a specific demographic models that includes all data, but we will typically still be able to explore such data e.g. using PCA. One way to justify $F$-statistics is to interpret the populations $X_i$ geometrically as points or vectors in the $S$-dimensionsl \emph{data space} $\mathbb{R}^S$.  \cite{oteo-garcia2021} showed that in this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space. In particular the $F$-statistics can be defined as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. \cite{oteo-garcia2021} showed that in this framework, $F$-statistics are closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.

The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions. However, it has been commonly observed that population structure is quite low-dimensional, with the first few PCs providing a good approximation of the underlying population structure. Therefore, we may hope that PCA could yield a reasonable approximation of the data space.

\begin{figure}[!ht]
	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}}
	\label{fig:outgroupf3}
\end{figure}

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variation of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics-framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have genotype data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way one can define PCA is that it aims to find a $K$-dimensional subspace of the data space that retains the most important variation. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	\begin{equation}
	\MY = \MC\MX = (\mathbf{U} \MSINGULAR) \mathbf{V}^T = \MP\ML\text{,} \label{eq:svd}
	\end{equation}

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure. The SNP loadings $\ML=\mathbf{V}^T$ form an orthonormal basis of size $n \times S$, its rows give the contribution of each SNP to each PC. It is often used to look for outliers, which might be indicative of selection \citep[e.g][]{duforet-frebourg2016}. Alternatively, the PCs can also be motivated as an eigendecomposition of the covariance matrix $\MY\MY^T$. This can be motivated from (\ref{eq:svd}):

	\begin{equation}
	\MY\MY^T = \MP\ML\ML^T\MP^T = \MP\MP^T,
	\end{equation}
	since $\ML\ML^T=\mathbf{I}$.

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

\subsubsection{$F$-statistics and projection on PCA}
One of the easiest ways of dealing with missing data in PCA is to calculate the principal components (equation \ref{eq:svd}) only on a subset of the data with no missingness, and then to \emph{project} the lower quality samples with high missingness onto this PCA. The simplest way to do this is to note that

\begin{equation*}
\MY\ML^T = \MP\ML\ML^T = \MP,
\end{equation*}
and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PC simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC. However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original sample. In this case, there is a projection error 
$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML }.$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.

%Equivalently, if we have $F_2$-distances from the 
%$$p_{\text{new}} = \frac{1}{2} \MSINGULAR^{-2}P^T d,$$
%where $d_i = \normsq{X_i} - F_2(X_i, X_{\text{new}})$ \citep{gower1968}. In this case,

Thus, if we compare the $F$-statistic of a projected sample, we have

\begin{equation}
F_2(X_i, X_{\text{new}}) = F_2(Y_i, Y_{\text{new}}) = F_2(P_i, P_{\text{proj}}) + F_2(P_{\text{proj}}\ML, Y_{\text{new}}),
\end{equation}
because the projection error and projection are orthogonal to each other.

@@

\subsection{When are admixture-$F_3$ statistics negative?}
Consider the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to the sampled population $X_x$. How is such a scenario displayed on a PCA?  ince the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B). In fact, if $X_x$ (and other populations related to $X_x$) was not part of the construction of the PCA, and is instead projected onto it, the drift from $X_y$ to $X_x$ will be independent of the population on the PCA, and $X_x$ will be at the same location on a PCA plot as $X_y$, which can be used to predict the admixture proportions \citep{brisbin2012, mcvean2009, oteo-garcia2021}.

If $X_x$, or populations more closely related to $X_x$ than $X_2$ and $X_3$, are included in the construction of the PCA, this is no longer true, and $X_x$ will project on a different spot than $X_y$ (Figure \ref{fig:admix}B). Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them?

One way to address this question is to consider the space for which $F_3$ is negative, i.e. 
\begin{eqnarray}
2 F_3(X_x; X_2, X_3) &=& 2\langle  X_x - X_2, X_x - X_3 \rangle \nonumber\\
      &=& \normsq{X_x - X_2} + \normsq{X_x - X_3}  - \normsq{X_2 - X_3} < 0 \label{eq:f3neg}\text{.}
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_2, X_3$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_2X_3}$ (The overline denotes a line segment). $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).

\paragraph{$F_3$ on a 2D PCA-plot.} 
 If we project this $n$-ball on a two-dimensional plot, $\overline{X_2X_3}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot. This geometry is perhaps easiest visualized on a globe. If we look at the globe from a view point parallel to the equator, both the north and south poles are visible at the very edge of the circle. But if we look at it from above the north pole, the north- and south-poles will be at the very same point.

@@

\paragraph{Example}
As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g. Sardinian, Finnish) have, in fact, positive $F_3$-values. This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.  Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle. This allows us to immediately conclude that their $F_3$-statistics must be positive; and we should not consider them as a mixture between Basques and Turks.

@@

\subsection{$F_3$-statistics as projections}
Outgorup-$F_3$-statistic motivate a comparison of $F_3$-statistics to projections (Equation \ref{eq:proj}), consider again the case displayed in Figure \ref{fig:outgroupf3}A, where the goal is to find the population $X_i$ that is closest to an unknown population $X_U$, with respect to an outgroup $X_O$, which we can do using the statistic $F_3(X_O; X_U, X_i)$. On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
$$\vectorproj[X_U - X_O]{X_i - X_O} = \frac{F_3(X_O; X_U, X_i)}{F_2(X_O; X_U)}(X_U - X_O).$$

Of the right-hand-side terms, $X_U-X_O$ gives the direction of the resulting vector, and the $F_2$-term in the denominator is a normalizing constant. Neiter of these terms depend on $X_i$, which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly. Thus, the outgroup-$F_3$-statistic is larger for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
